An enhancer code compendium of early human neuronal and neural crest development

### 2025-04-08 manuscript.tex
M&M V.C.

Methods details below. All human iPSC and ESC experiments were approved by the Ethical Committee of the KU Leuven  Number
%TODO: add ECD

@@

\SI{1}{\milli\litre} pipette. Immediately after the TrypLE reaction was inhibited by adding \SI{2}{\milli\liter} of DMEM/F12 FBS (Fetal Bovine Serum (FBS; ThermoFisher, A5256701) \SI{20}{\percent} V/V; Penicillin-Streptomycin (ThermoFisher, 15140122) \SI{50}{\micro\gram} in DMEM/F12 (ThermoFisher, 11320033)). Cells were pelleted at 300 RCF for \SI{3}{\minute} and after counting on a LUNA-FL™ Dual Fluorescence Cell Counter resuspended to \SI{3.5e6}{cells\per\milli\litre} in $GM_0$ ( basal GM (N-2 Supplement (ThermoFisher, 17502048) 1x, B-27 Supplement (ThermoFisher, 17504044) 1x, GlutaMax Supplement (ThermoFisher, 35050061) 1x, Sodium Pyruvate (ThermoFisher, 11360070) \SI{1}{\milli\molar}, MEM Non-Essential Amino Acids Solution (ThermoFisher, 11140050) 1x, Penicillin-Streptomycin (ThermoFisher, 15140122) \SI{50}{\micro\gram\per\milli\litre}, Neurobasal Medium (ThermoFisher, 21103049) \SI{50}{\percent} V/V and DMEM/F12 (ThermoFisher, 11320033) \SI{50}{\percent} V/V) with RHO/ROCK inhibitor (Y-27632 (Dihydrochloride), Stemcell technologies, 100-1044) \SI{1}{\micro\molar}). Next, the polyethylene glycol A (PEG-A) hydrogel matrix (see Methods detail below) was prepared by first adding Corning Mouse Laminin (Avantor, 734-1098) \SI{0.1}{\milli\gram\per\milli\liter} followed by activated factor 13 (see Methods details below) in that order to a PEG mixture (see Methods details below; PEG \SI{1.5}{\percent} or \SI{2}{\percent} V/V, buffer (TRIS \SI{0.05}{\milli\molar}, $CaCl_2$ \SI{0.05}{\milli\molar}; pH 7.6; filter sterilized) in $GM_0$ medium). After mixing the matrix by pipetting up and down, without generating bubbles, for 1-2 minutes \SI{3.5e4}{cells\per\micro\litre} mixture were added. The matrix and gell mixture was mixed gently one more time by pipetting and \SI{10}{\micro\litre} droplets were generated in each well of a TC-treated 96 well plate (Corning, 3603). When the matrix started to gell the plate was turned up-side down for 10 seconds, after which it was turned right-side up for 10 seconds. This was repeated for a total of 5 times, after which the plates were kept at room temperature. After 20 minutes of gelling \SI{200}{\micro\litre} of $GM_0$ medium was added to each well and the organoids were incubated at \SI{37}{\degreeCelsius} \SI{5}{\percent} $CO_2$ for a total of 11 days, or until the timepoint of single-cell multiome. At day 3 of differentiation medium was changed with basal GM supplemented with retinoic acid (Sigma-Aldrich, R2625-100MG) \SI{0.25}{\micro\molar} and Smoothened agonist (Sigma-Aldrich, 566661-500UG) \SI{1}{\micro\molar}. At day 5, 7 and 9 the medium was changed with basal GM.

\subsubsection*{Chicken embryos}
Fertilized chicken eggs, obtained from ... 
%TODO: add vendor
were incubated at \SI{37}{\degreeCelsius} at \SI{70}{\percent} humidity for 30 hours to reach Humburger Hamilton (HH) stage 4. Afterwards, embryos were removed from the egg and kept in culture for an additional 24 hours.

@@

\paragraph*{single cell suspension}
PEG hydrogel droplets were removed from the 96 well plate by gently scraping with a \SI{100}{\micro\litre} pipette tip
and transferred to a tube on ice using a p1000 pipette by putting the pipette tip on the droplet and creating a gentle
negative pressure, making sure that the droplet does not enter the pipette tip. Organoids were dissociated by adding room temperature TrypLE Express Enzyme (ThermoFisher, 12605010) and incubation at \SI{37}{\degreeCelsius} while pipetting up and down until no large pieces of hydrogel were visible anymore or a maximum of 7.5 minutes. Immediately after the TrypLE reaction was inhibited by adding twice the volume of DMEM/F12 FBS (Fetal Bovine Serum (FBS; ThermoFisher, A5256701) \SI{20}{\percent} V/V; Penicillin-Streptomycin (ThermoFisher, 15140122) \SI{50}{\micro\gram} in DMEM/F12 (ThermoFisher, 11320033)). The single-cell suspension was pelleted at 500 RCF for 4 minutes, resuspended in DMEM/F12 FBS and counted on LUNA-FL™ Dual Fluorescence Cell Counter. Immediately after the cells were processed for 10x single-cell multiome.

@@

\paragraph*{10x multiome library preparation}
%TODO: Valerie

\paragraph*{10x multiome library sequencing}
%TODO: Valerie

>>>>

Methods details below. 
All human iPSC and ESC experiments were approved by the Ethical Committee of the KU Leuven (ECD S63316)

@@

\SI{1}{\milli\litre} pipette. Immediately after the TrypLE reaction was inhibited by adding \SI{2}{\milli\liter} of DMEM/F12 FBS (Fetal Bovine Serum (FBS; ThermoFisher, A5256701) \SI{20}{\percent} V/V; Penicillin-Streptomycin (ThermoFisher, 15140122) \SI{50}{\micro\gram} in DMEM/F12 (ThermoFisher, 11320033)). Cells were pelleted at 300 RCF for \SI{3}{\minute} and after counting on a LUNA-FL Dual Fluorescence Cell Counter resuspended to \SI{3.5e6}{cells\per\milli\litre} in $GM_0$ ( basal GM (N-2 Supplement (ThermoFisher, 17502048) 1x, B-27 Supplement (ThermoFisher, 17504044) 1x, GlutaMax Supplement (ThermoFisher, 35050061) 1x, Sodium Pyruvate (ThermoFisher, 11360070) \SI{1}{\milli\molar}, MEM Non-Essential Amino Acids Solution (ThermoFisher, 11140050) 1x, Penicillin-Streptomycin (ThermoFisher, 15140122) \SI{50}{\micro\gram\per\milli\litre}, Neurobasal Medium (ThermoFisher, 21103049) \SI{50}{\percent} V/V and DMEM/F12 (ThermoFisher, 11320033) \SI{50}{\percent} V/V) with RHO/ROCK inhibitor (Y-27632 (Dihydrochloride), Stemcell technologies, 100-1044) \SI{1}{\micro\molar}). Next, the polyethylene glycol A (PEG-A) hydrogel matrix (see Methods detail below) was prepared by first adding Corning Mouse Laminin (Avantor, 734-1098) \SI{0.1}{\milli\gram\per\milli\liter} followed by activated factor 13 (see Methods details below) in that order to a PEG mixture (see Methods details below; PEG \SI{1.5}{\percent} or \SI{2}{\percent} V/V, buffer (TRIS \SI{0.05}{\milli\molar}, $CaCl_2$ \SI{0.05}{\milli\molar}; pH 7.6; filter sterilized) in $GM_0$ medium). After mixing the matrix by pipetting up and down, without generating bubbles, for 1-2 minutes \SI{3.5e4}{cells\per\micro\litre} mixture were added. The matrix and gell mixture was mixed gently one more time by pipetting and \SI{10}{\micro\litre} droplets were generated in each well of a TC-treated 96 well plate (Corning, 3603). When the matrix started to gell the plate was turned up-side down for 10 seconds, after which it was turned right-side up for 10 seconds. This was repeated for a total of 5 times, after which the plates were kept at room temperature. After 20 minutes of gelling \SI{200}{\micro\litre} of $GM_0$ medium was added to each well and the organoids were incubated at \SI{37}{\degreeCelsius} \SI{5}{\percent} $CO_2$ for a total of 11 days, or until the timepoint of single-cell multiome. At day 3 of differentiation medium was changed with basal GM supplemented with retinoic acid (Sigma-Aldrich, R2625-100MG) \SI{0.25}{\micro\molar} and Smoothened agonist (Sigma-Aldrich, 566661-500UG) \SI{1}{\micro\molar}. At day 5, 7 and 9 the medium was changed with basal GM.

\subsubsection*{Chicken embryos}
Fertilized chicken eggs, obtained from "Kwekerij van het Hallerbos" were incubated at \SI{37}{\degreeCelsius} at \SI{70}{\percent} humidity for 30 hours to reach Humburger Hamilton (HH) stage 4. Afterwards, embryos were removed from the egg and kept in culture for an additional 24 hours.

@@

\paragraph*{single cell suspension}
PEG hydrogel droplets were removed from the 96 well plate by gently scraping with a \SI{100}{\micro\litre} pipette tip
and transferred to a tube on ice using a p1000 pipette by putting the pipette tip on the droplet and creating a gentle
negative pressure, making sure that the droplet does not enter the pipette tip. Organoids were dissociated by adding room temperature TrypLE Express Enzyme (ThermoFisher, 12605010) and incubation at \SI{37}{\degreeCelsius} while pipetting up and down until no large pieces of hydrogel were visible anymore or a maximum of 7.5 minutes. Immediately after the TrypLE reaction was inhibited by adding twice the volume of DMEM/F12 FBS (Fetal Bovine Serum (FBS; ThermoFisher, A5256701) \SI{20}{\percent} V/V; Penicillin-Streptomycin (ThermoFisher, 15140122) \SI{50}{\micro\gram} in DMEM/F12 (ThermoFisher, 11320033)). The single-cell suspension was pelleted at 500 RCF for 4 minutes, resuspended in DMEM/F12 FBS and counted on LUNA-FL Dual Fluorescence Cell Counter. Immediately after the cells were processed for 10x single-cell multiome.

@@

\paragraph*{10x multiome nuclei isolation}
To isolate nuclei from the single cells of the dissociated neural tube organoids, protocol CG000365 (10x Genomics) 
was followed. 
For each developmental time point, 500k cells were resuspended in 100 µl nuclei lysis buffer 
(\SI{10}{\milli\molar} Tris-HCl pH 7.4; 
\SI{10}{\milli\molar} NaCl; 
\SI{3}{\milli\molar} $MgCl_2$; 
\SI{0.1}{\percent} Tween-20; 
\SI{0.1}{\percent} NP40; 
\SI{0.01}{\percent} Digitonin;
\SI{1}{\percent} BSA; 
\SI{1}{\milli\molar} DTT and 
\SI{1}{\units\per\micro\litre} RNasin ribonuclease inhibitor 
in nuclease-free water
). 
After 5 min incubation on ice, \SI{1}{\milli\litre} chilled wash buffer was added to the lysed cells 
(
\SI{10}{\milli\molar} Tris-HCl pH 7.4; 
\SI{10}{\milli\molar} NaCl; 
\SI{3}{\milli\molar} $MgCl_2$; 
\SI{0.1}{\percent} Tween-20; 
\SI{1}{\percent} BSA; 
\SI{1}{\milli\molar} DTT and 
\SI{1}{\units\per\micro\litre} RNasin ribonuclease inhibitor in 
nuclease-free water
). 
Nuclei were pelleted by centrifugation at 500 g for 5 min at \SI{4}{\degreeCelsius} and 
resuspended in a 1x nuclei buffer (10x Genomics) supplemented with 
\SI{1}{\milli\molar} dithiothreitol and \SI{1}{\units\per\micro\litre} RNasin ribonuclease inhibitor. 
Nuclei suspensions were passed through a \SI{40}{\micro\meter} Flowmi filter (VWR Bel-Art SP Scienceware). 
Nuclei concentration was assessed using the LUNA-FL Dual Fluorescence Cell Counter.

\paragraph*{10x multiome library preparation}
Single-cell libraries were generated using the 10X Chromium Single-Cell Instrument and NextGEM Single Cell Multiome ATAC+Gene Expression kit (10X Genomics) according to the manufacturer’s protocol. 
In brief, single nuclei of neural tube organoids were incubated for 60 min at \SI{37}{\degreeCelsius} 
with a transposase that fragments the DNA in open regions of the chromatin and adds adapter sequences to the 
ends of the DNA fragments. 
After generation of nanoliter-scale gel bead-in-emulsions (GEMs), 
GEMs were incubated in a C1000 Touch Thermal Cycler (Bio-Rad) under the following program: 
\SI{37}{\degreeCelsius} for 45 min, \SI{25}{\degree} for 30 min, and hold at \SI{4}{\degreeCelsius}. 
Incubation of the GEMs produced 10x barcoded DNA from the transposed DNA (for ATAC) and 10x barcoded, 
full-length cDNA from poly-adenylated mRNA (for GEX). 
This was followed by a quenching step that stopped the reaction. 
After quenching, single-cell droplets were broken and the transposed DNA and full-length cDNA were isolated using
Cleanup Mix containing Silane Dynabeads. 
To fill gaps and generate sufficient mass for library construction, the transposed DNA and cDNA were amplified via PCR: 
\SI{72}{\degreeCelsius} for 5 min; 
\SI{98}{\degreeCelsius} for 3 min; 
seven cycles of 
\SI{98}{\degreeCelsius} for 20 s, 
\SI{63}{\degreeCelsius} for 30 s, 
\SI{72}{\degreeCelsius} for 1 min; 
\SI{72}{\degreeCelsius} for 1 min; 
and hold at \SI{4}{\degreeCelsius}. 
The pre-amplified product was used as input for both ATAC library construction and cDNA amplification for gene
expression library construction. 
Illumina P7 sequence and a sample index were added to the single-strand DNA during ATAC library construction via PCR:
\SI{98}{\degreeCelsius} for 45 s;
7–9 cycles of \SI{98}{\degreeCelsius} for 20 s, 
\SI{67}{\degreeCelsius} for 30 s, 
\SI{72}{\degreeCelsius} for 20 s;
\SI{72}{\degreeCelsius} for 1 min; 
and hold at \SI{4}{\degreeCelsius}. 
The sequencing-ready ATAC library was cleaned up with SPRIselect beads (Beckman Coulter). 
Barcoded, full-length pre-amplified cDNA was further amplified via PCR: 
\SI{98}{\degreeCelsius} for 3 min; 
6–9 cycles of 
\SI{98}{\degreeCelsius} for 15 s, 
\SI{63}{\degreeCelsius} for 20 s, 
\SI{72}{\degreeCelsius} for 1 min; 
\SI{72}{\degreeCelsius} for 1 min; 
and hold at \SI{4}{\degreeCelsius}. 
Subsequently, the amplified cDNA was fragmented, end-repaired, A-tailed, and index adaptor ligated, 
with SPRIselect cleanup in between steps. 
The final gene expression library was amplified by PCR: 
\SI{98}{\degreeCelsius} for 45 s; 
5–16 cycles of 
\SI{98}{\degreeCelsius} for 20 s,
\SI{54}{\degreeCelsius} for 30 s, 
\SI{72}{\degreeCelsius} for 20 s, 
\SI{72}{\degreeCelsius} for 1 min; 
and hold at \SI{4}{\degreeCelsius}. 
The sequencing-ready GEX library was cleaned up with SPRIselect beads.

\paragraph*{10x multiome sequencing}
Before sequencing, the fragment size of every library was analyzed using the Bioanalyzer high-sensitivity chip. 
All 10x single cell Multiome ATAC libraries were sequenced on NextSeq2000 instrument (Illumina) or on NovaSeq6000
instrument (Illumina) 
with the following sequencing parameters: 
51 bp read 1, 
8 bp index 1, 
24 bp index 2, 
51 bp read 2. 
All 10x single cell Multiome Gene Expression libraries were sequenced on NextSeq2000 instrument (Illumina) or on NovaSeq6000 instrument (Illumina) 
with the following sequencing parameters: 
28 bp read 1, 
10 bp index 1, 
10 bp index 2, 
90 bp read 2.
